Reconstructing the last known movements of one of Nature's giants

### 2018-05-02 manuscript/CurrentBiology/blue-whale-ms.tex
Edits to intro

\newpage
\section{Introduction}
Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}. 
Identifying threats to migratory species and developing effective conservation measures requires information on individual-level movements spanning multiple years, for both historical and present-day populations. 
Such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Despite hunting for blue whales being banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List \cite{reilly2008balaenoptera}.
In the northeast Atlantic, blue whales appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, now fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
Following movements of individual blue whales is extremely challenging due to their scarcity and oceanic habit.
Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggest that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde islands \cite{baines2014upwellings}.
Northward migration in the spring may occur along mid-Atlantic corridors with peak sightings in the Azores around April-May. Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}.
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southward-migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
While direct tagging methods are adding vital behavioural data to our understanding of whale movements \cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
Conservation efforts are consequently hampered by the lack of detailed data on the distribution, behaviour and ecology of blue whales, particularly during the winter and for the juvenile phase of life. 
An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotopes \cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
Stable isotope data therefore provide information on diet and location (i.e. `trophic geography' \cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values. 
Here we reconstruct movements of an historic whale by comparing isotopic compositions of samples from museum collections, rich archives of behavioural information \cite{lister2011natural}, to spatially referenced models of isotopic variation in diet across the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}. 
Mysticete (baleen) whales are characterized by the development of baleen plates, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \cite{best1996stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, an individual's entire lifespan is rarely recorded. 
Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales \cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population. 
Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope'', a female whale estimated to be a young adult c.15 years at her death in 1891.

@@

Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on.
We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope \cite{west2006stable}.
Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information \cite{west2006stable}.

Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplemental Methods; Supplemental Figures S1 and S2). 
Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (from the start of the record in early 1884 to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplemental Figures S1 and S2). 
In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively high and $\delta^{13}$C values are relatively low with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively high $\delta^{13}$C values in the second half of 1891. 
The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplemental Figures S1 and S2).

>>>>

\newpage
\section{Introduction}
Migratory species pose a particular challenge for conservation practitioners because their effective conservation relies on protection at many, often distant, sites \cite{runge2014conserving}.
Migratory species may also be particularly vulnerable to changes in climate or human use of the environment, as they are influenced by conditions in multiple locations across different parts of their life cycle \cite{robinson2009travelling}. 
Identifying threats to migratory species, understanding species responses to change, and developing effective conservation measures all require information on individual-level movements spanning multiple years, ideally for both historical and present-day populations.
With the development of electronic tagging technology, studies of animal populations have shifted from infrequent observations of many individuals with limited or no information about individual movement behavior to frequent observations of a few individuals, with detailed information about individual movement \cite{holdo2013inferring}. 
However, such data are scarce, especially for rare, wide-ranging and long-lived marine species \cite{ryan2013stable,hall2005stable,bailey2009behavioural}.

The blue whale (\textit{Balaenoptera musculus}), the largest animal to have ever lived, is one such migratory marine species. 
Although hunting for blue whales has been banned for over 50 years, blue whales are still listed as Endangered on the IUCN Red List \cite{reilly2008balaenoptera}.
Blue whales in the northeast Atlantic appear to form a separate population, with little evidence for exchange between northeast and northwest Atlantic populations \cite{pike2009note}.  
The northeast Atlantic population of blue whales was the first large whale population to be systematically hunted with explosive harpoons, and while the population was probably relatively small before hunting, currently fewer than 1000 individuals are estimated to be in the northeast Atlantic in summer, mainly distributed around Iceland \cite{pike2009note}.
Their small population size and oceanic habit makes blue whale movements extremely hard to study.

Current methods for tracking large whales include satellite tracking, radio tagging, aerial photography, passive acoustic monitoring, and sightings records. 
While these methods are adding vital behavioural data to our understanding of whale movements \cite{borger15,mcdonald2006biogeographic,bailey2009behavioural,mate2007evolution,silva2013north}, they are often expensive, limited to short-term studies \cite{bailey2009behavioural,best2015tag,mate2007evolution}, and may influence whale behaviour \cite{walker2012review}.
At present the longest period of continuous monitoring via satellite tagging of an individual blue whale in the north Atlantic, that we are aware of, is 177 days \cite{lesage2017foraging}, but most satellite deployments on blue whales report movement data for under three months \cite{heide2001new,silva2013north, bailey2009behavioural,lesage2017foraging}, and are unable to show individual movements between feeding and wintering areas. 
To our knowledge photographic identification has matched summer and winter locations of a single north-east Atlantic blue whale with sightings in Iceland and Mauritania (REF).

Similar to other large balaenopterid whales, blue whales in the northeast Atlantic are generally assumed to conduct annual migrations between high and low latitudes. 
A combination of historic whaling data, observations, satellite tracks and acoustic monitoring suggests that blue whales in the northeast Atlantic track regions of high production throughout the year, with at least some whales wintering in the upwelling systems between Mauritania and the Cape Verde Islands \cite{baines2014upwellings}. 
Northward migration in the spring may occur along mid Atlantic corridors with peak sightings in the Azores around April-May \cite{silva2013north}. 
Summer feeding appears to occur in northerly latitudes around Iceland, and historically in the Norwegian and Barents Seas \cite{pike2009note}. 
Blue whales are frequently detected in waters to the west of the UK, with peak acoustic detections occurring between November and December, in southerly migrating animals \cite{reeves2004historical,baines2017autumn,charif2009acoustic,visser2011timing}.
Inferences concerning movements of blue whales in the northeast Atlantic are therefore largely drawn from disparate information sources and rarely detail movements of individual animals. 
It is difficult to know whether spatial variations in observed distributions reflect movements undertaken by individual animals or the frequency that individuals may undertake movements. 
Information on historical movement patterns of individual blue whales is particularly important for understanding the drivers of whale declines, and potential human-whale conflicts as populations recover.

An alternative technique for investigating whale movements retrospectively is to use intrinsic biochemical information such as stable isotope compositions \cite{west2006stable,busquets2017estimating,hobson2008tracking}. 
Present-day populations can be studied with material collected in the field, while historic samples can be taken from museum collections; rich archives of behavioral information that are often under-utilised \cite{lister2011natural}. 
The stable isotope composition of animal tissues reflects the isotopic composition of diet at the time and place of ingestion, integrated over the timescale of tissue growth. 
Stable isotope data therefore provide information on diet and location (i.e. `trophic geography' \cite{bird2018global}), but incomplete knowledge of the spatio-temporal distribution of the isotopic composition of prey currently limits interpretation of measured values. 
Here we use the isotopic compositions of whale tissues to reconstruct movement by comparing them to spatially referenced models of isotopic variation in diet across the species' range in the northeast Atlantic \cite{hobson1999tracing,hobson2008tracking,eisenmann2016isotopic}.

Mysticete (baleen) whales are characterized by the development of baleen, keratinous structures in the upper jaw used to filter food items from seawater. 
Baleen is ideal for stable isotope studies because keratin grows continuously through an individual's life, and once laid down it is metabolically inert \cite{best1996stable,hobson1998stable}. 
Baleen plates therefore offer a continuous isotopic record of behavior reflecting multiple years of life of an individual whale, although as baleen is worn away at the tips over time, so rarely records an individual's entire lifespan. 
Baleen stable isotope records have a long history of use as biochemical markers to infer movement patterns in whales \cite{ryan2013stable,best1996stable,hobson1998stable,hobson2008tracking}, but have yet to be applied to blue whales from the North Atlantic population. 
Here we apply this method to baleen from the Natural History Museum London's (NHM) blue whale ``Hope''. 
Hope stranded in Wexford Ireland in March 1891 and is a young adult female, estimated to be at least 15 years at death, suggesting that her baleen will yield information about migratory behaviors in north Atlantic blue whales around the peak of industrial whaling.

@@

Blue whales feed mainly on krill (Euphausiacea), small crustaceans that feed on phytoplankton \cite{handbook}.
Thus the stable isotope compositions of keratin at a given point in the baleen will reflect the stable isotope compositions of the krill it was feeding on in the weeks prior to keratin growth.
We sampled keratin from the blue whale's baleen plate at 1cm intervals perpendicular to the direction of growth, resulting in 97 discrete samples of baleen.
We then used continuous-flow isotope ratio mass spectrometry to determine the isotopic compositions of carbon and nitrogen in the keratin.
These isotopes are expressed as $\delta^{13}$C and $\delta^{15}$N values, i.e. relative to a standard and as ratios of the rare ($^{13}$C, $^{15}$N) to common ($^{12}$C, $^{14}$N) isotope \cite{west2006stable}.
Carbon and nitrogen isotopes are commonly reported together as they provide complementary information; both isotopes provide information on the geographical locations of an individual, but nitrogen isotopes are more strongly fractionated during protein metabolism and thus contain additional trophic and physiological information \cite{west2006stable}.

Throughout the record, $\delta^{15}$N values record regular cyclical fluctuations, with a mean spacing of 13.5cm, assumed to represent annual periodicity (see Supplemental Methods; Supplemental Figures S1 and S2). 
Therefore, given the date of stranding (25th March 1891), and estimated baleen growth rates of 13.5cm $y^{-1}$, we reconstructed a timeline for $\delta^{13}$C and $\delta^{15}$N fluctuations in the baleen over seven full years of the whale's life (early 1884 - spring 1891). 
The $\delta^{13}$C and $\delta^{15}$N profiles show two distinct phases in the whale's behaviour. 
In behavioural phase one (from the start of the record to spring 1886), we find relatively stable, elevated $\delta^{13}$C values, and relatively low $\delta^{15}$N values (Figure \ref{fig1}; Supplemental Figures S1 and S2). 
In behavioural phase two (summer 1886 to spring 1890) $\delta^{15}$N values are relatively high and $\delta^{13}$C values are relatively low with coincident cyclical fluctuations in both $\delta^{13}$C and $\delta^{15}$N values. 
In the last year of life the cyclical pattern is disrupted, with constant low $\delta^{13}$C values for approximately six months in the first half of 1890, before a rapid switch to relatively high $\delta^{13}$C values in the second half of 1890. 
The final three months of the record show a progressive fall in $\delta^{13}$C values (Figure \ref{fig1}; Supplemental Figures S1 and S2).
